---
output:
  pdf_document: default
  html_document: default
---
# Random variables 01

##### Josep Fortiana 2018-09-26

Load auxiliary functions

```{r}
source( "Random.vars.r" )
```

# Discrete random variables

## A discrete r.v.

```{r}
x<-c(1,2,3,4)
d<-c(0.2, 0.3, 0.4, 0.1)
```

#### Probability mass function (pmf)

```{r}
plotmass(x,d)
```

#### A visualization of the probability distribution as a set of probability mass points on the line

```{r}
plotmasspoints(x,d)
```

#### The cumulative distribution function (cdf) or probability distribution function

```{r}
plotdist(x,d)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Constant (degenerate) r.v.

#### Vector of values, $x$, and  vector of probabilities, $d$

Both have a single element each

```{r}
x<-2
d<-1
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Bernoulli r.v.

```{r}
x<-c(0,1)
d<-c(0.2,0.8)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Hypergeometric r.v.

```{r}
# Results of drawing k=4 balls without replacement from an urn containing m=6 balls, n=4 of which are white 
#(and m-n=2 are black). The hypergeometric random variable $X$ is the number of white balls obtained.
m<-6
n<-4
k<-4
# Vector of values
x<-0:4
# Vector of probabilities
d<-dhyper(x, m, n, k)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Binomial r.v.

```{r}
# Binomial B(6,0.4) r.v.
n<-6
p<-0.4
x<-0:n
```

#### Joint plot of pmf and cdf

```{r}
d<-dbinom(x, size=n,prob=p)
plotmassdist(x,d)
```

```{r}
# Binomial B(10,0.5) r.v.
n<-10
p<-0.5
x<-0:n
d<-dbinom(x, size=n,prob=p)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Geometric r.v.

```{r}
# Geometric Geom(0.4) r.v.
n<-10
p<-0.4
x<-0:n
d<-dgeom(x,prob=p)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Negative binomial r.v.

```{r}
p<-0.4
r<-4
y<-0:16  
d<-dnbinom(y,size=r,prob=p)
```

#### Probability mass function (pmf)

```{r}
plotmass(y,d)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(y,d)
```

## Poisson r.v.

```{r}
# Poisson Pois(2) r.v.
n<-8
lambda<-2
x<-0:n
d<-dpois(x,lambda=lambda)
```

#### Probability mass function (pmf)

```{r}
plotmass(x,d)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

```{r}
# Poisson Pois(5) r.v.
lambda<-5
n<-15  # A sensible maximum x for plotting (arrived at by trial and error)
x<-0:n
d<-dpois(x,lambda=lambda)
```

#### Probability mass function (pmf)

```{r}
plotmass(x,d)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

## Discrete uniform r.v. (generalized die)

```{r}
# Discrete uniform r.v. (generalized die with n=9 faces)
n<-9
x<-1:n
d<-rep(1/n,n)
```

#### Joint plot of pmf and cdf

```{r}
plotmassdist(x,d)
```

# Continuous random variables

## Uniform r.v. on a (compact) interval

#### Probability density function (pdf) and cumulative distribution function (cdf)

The shape of this density gives the distribution its name as "rectangular distribution"

```{r}
# Uniform distribution on [0,2]
f<-function(x){dunif(x,0,2)}
F<-function(x){punif(x,0,2)}
```

#### Joint plot of pdf and cdf

```{r}
xmin<--1
xmax<-3
plotdensdist(f,F,xmin,xmax)
```

## Exponential r.v.

#### Probability density function (pdf) and cumulative distribution function (cdf)

```{r}
# Exponential distribution with lambda=2
f<-function(x){dexp(x,rate=2)}
F<-function(x){pexp(x,rate=2)}
```

#### Joint plot of pdf and cdf

```{r}
xmin<--1
xmax<-6
plotdensdist(f,F,xmin,xmax)
```

```{r}
```